\large How resource abundance and stochasticity affect organisms' space-use requirements

### 2024-02-02 writing/manuscript.Rmd
started major edits

---
title: \large How resource abundance and stochasticity affect organisms' space-use requirements
# author:
#   - name: Stefano Mezzini
#     institute: [braes, biol]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [braes, biol, cmps]
#     correspondence: "yes"
# institute:
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl' # use 'nature.csl' to count refs
# biblio-style: 'amnatnat'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has character U+0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

<!-- start main body on a new page -->
\newpage

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018], along with changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018]. Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020]. However, movement represents one of the most readily available behaviors that species can adjust, whether this be by modifying their home range [@hirt_environmental_2021; @bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].

@@

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020]. While many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, few studies build upon others' previous work. Instead, results are typically reported as independent, novel findings. Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' space-use requirements. In addition, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.

Here, we refer to a location's average amount of resources as resource abundance, while we use the phrase resource stochasticity to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for. Although it is generally expected that resource unpredictability will decrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

@@

\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the amount of resources at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.

In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically, since resources tend to be less variable when they are less abundant. We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix A.

@@

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982]. The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see fig. \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].

@@

When $\e(R)$ is high, we expect organisms' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. However, as $\e(R)$ decreases, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organisms' space-use requirements and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an organism's lifespan. For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access. Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022]. Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations. Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in fig. \@ref(fig:mu-hyp).

Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see fig. A1). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (fig. A2). Since $R$ is spatiotemporally random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict where $R$ is high if $\e(R)$ does not vary predictably over space.

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case. Most often, $\e(R)$ will vary over time, and organisms' behaviors will change in response.

@@

\noindent Although the effect of $R$ on organisms' space-use requirements is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\var(R)$ can fluctuate temporally (and spatially) due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones. For example, since berries are scarce outside the growing season, the variance in berries will also be low. But during the growing season, both the average number of berries and the variance in berries are higher, since a bear may explore a location before fruiting time or after competitors ate all the berries. In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].

@@

@@

# Applying the framework

@@

>>>>

---
title: \large How resource abundance and stochasticity affect organisms' space-use requirements
# author:
#   - name: Stefano Mezzini
#     institute: [braes, biol]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [braes, biol, cmps]
#     correspondence: "yes"
# institute:
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl' # use 'nature.csl' to count refs
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has character U+0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

<!-- start main body on a new page -->
# To do:

* make sure appendices flow (added stuff from main text)

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018], along with changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018]. Thus, many organisms adapt their behaviors in response to changes in local resource abundance to ensure their needs are met.<!-- Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020]. However, --> Movement represents one of the most readily available behaviors that species can adjust to optimize their resource use [<!--@krivan_ideal_2008; @brown_ecology_1999; -->@kacelnik_ideal_1992; @charnov_optimal_1976].<!-- whether this be by modifying their home range [@hirt_environmental_2021; @bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].-->

@@

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but it fails to account for the nonlinear relationship required to avoid negative home range estimates [e.g., Figure 2 in the work of @bista_effect_2022].<!-- to an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].--> Many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, but studies often fail to build upon previous work. Instead, results are typically reported as independent, novel findings. Thus, there remains a need for a clear hypothesis for the effects of resource abundance on organisms' space-use requirements. Additionally, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources [but see: @stephens_optimal_1982; @rizzuto_forage_2021].

Here, we refer to a location's average amount of resources as "resource abundance", while we use the phrase "resource stochasticity" to indicate the variability in resources after accounting for changes in the mean. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. For instance, herbivores in a grassland with relatively low but constant forage availability (low mean and variance) will adopt different behaviors and adaptations from herbivores in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (low mean and high stochasticity). In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [@teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal in search for high-resource patches and short-term range residency whinith patches [@charnov_optimal_1976]. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in different selection of movement and life history strategies. Although resource unpredictability (and seasonality) may ecrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

@@

\noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, let the random variable $R$ indicate the amount of resources at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics. We assume readers are familiar with fundamental statistical concepts (e.g., random variables, probability distributions, expected value and variance of a random variable), but we define and explain the necessary concepts in Appendix A.

<!--
\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the amount of resources at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.

In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically, since resources tend to be less variable when they are less abundant. We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix A.
-->

@@

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982]. The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime and a minimum amount of space it requires to survive [which is necessarily greater than zero -- see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].

@@

Fig. \@ref(fig:mu-hyp) illustrates our first of two hypotheses: When $\e(R)$ is high, we expect organisms' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. At lower values of $\e(R)$, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organisms' space-use requirements and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, species-level changes in movement behavior are more likely to occur over evolutionary timescales than over an organism's lifespan, since larger ranges require greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

Consider the scenario where $\e(R) = \mu(t)$ changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (Fig. A1).<!-- We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant.--> Since $R$ is random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $\e(R)$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017] but will still not be able to predict $R$ exactly.

<!-- This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case. Most often, $\e(R)$ will vary over time, and organisms' behaviors will change in response. -->

@@

\noindent The example illustrated in the section above can occasionally be a useful simplification of relatively stable environments or when data is scarce, but it likely not realistic, since $\var(R)$ often also changes over time. Periods of scarcity (e.g., winter, droughts) not only have lower $\e(R)$ but also have narrower ranges of $R$. In contrast, periods of high $\e(R)$ (e.g., spring green-up) also often tend to have higher $\var(R)$ because $R$ tends to vary across wider ranges (e.g., due to high spatial variability, or high $R$ followed by low $R$ after grazing). When both $\e(R)$ and $\var(R)$ change over time (fig. A2), predicting $R$ becomes more complex, since disentangling changes in $\e(R)$ and $\var(R)$ is not simple [@steixner-kumar_strategies_2020]. Statistically, this is because the more change one attributes to $\mu(t)$ (i.e., the wigglier it is), the smaller $\sigma^2(t)$ becomes. Conversely, the smoother $\mu(t)$ is, the larger $\sigma^2(t)$ becomes. Biologically, this is important because an organism's perception scale [@levin_problem_1992] determines whether it sees a change in $R$ as a trend in $\e(R)$ or as a stochastic event (i.e., due to $\var(R)$). An organism's perception of changes in $R$ will also depend strongly on the its cognitive capacities and memory [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013]. Whether, an organism is able to predict trends in $\sigma^2(t)$ or not, environmental variability is thought to reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [e.g., @berger_climate_2018] and increases their space-use requirements. While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude [@logares_black_2012; @anderson_black-swan_2017]. A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still limited. Thus, there remains a need for developing a more complete understanding of how organisms' space-use requirements change with environmental stochasticity.

<!-- Although the effect of $R$ on organisms' space-use requirements is often recognized and accounted for in ecology [@stephens_optimal_1982; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\var(R)$ can fluctuate temporally (and spatially) due to many important drivers, including repetitive and predictable patterns. For example, since berries are scarce outside the growing season, the variance in berries will also be low. But during the growing season, both the average number of berries and the variance in berries are higher, since a bear may explore a location before fruiting time or after competitors ate all the berries. In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017]. -->

Similarly to $\e(R)$, we hypothesize $\var(R)$ to have a nonlinear effect on organisms' space-use requirements. When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect. But as resources become increasingly unpredictable, we expect space-use requirements to increase progressively faster (fig. \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases, the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], (3) stochasticity reduces an organism's ability to specialize and reduce competition for $R$ [@levins_evolution_1974], and (4) the added movement required to search for food increases organisms' energetic requirements. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching). If changes in $\sigma^2(t)$ are sufficiently predictable, organisms may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food, migrating, or relying on alternative food sources during stochastic events (e.g., the seasonal arrival of competitors).

@@

***HERE***

@@

***to do***:

* add a paragraph and figure on effects of $\mu(t)$ and $\sigma^2(t)$ on $\var(R)$.

# A lowland tapir in the Brazilian Cerrado: a case study

***to do***:

* open with limitations of sims
* then highlight shortcomings of empirical data
* move trasnformation stuff to the appendix
* move modeling details to appendix C

@@

Paragraphs to tie in:
